---
title: "team.Rmd"
author: "Caleb Purcell"
date: "9/28/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing Packages and Dataset (Please Excuse Warning Messages)

```{r}
library(tidyverse)
library(plyr)
library(dplyr)
library(ggplot2)
library(devtools)
library(readxl)
dat <- readxl::read_xls('~/ds202_lab3/GSS.xls')
```

## Manipulating Data Frame

```{r}
rename <- function(dat, oldnames, newnames) {
  datnames <- colnames(dat)
  datnames[which(datnames %in% oldnames)] <- newnames
  colnames(dat) <- datnames
  dat
}

dat <- rename(dat,c('Gss year for this respondent', 'Belief in life after death', 'Favor or oppose death penalty for murder', 'Political party affiliation', 'Respondents sex', 'Highest year of school completed', 'Marital status', 'Labor force status', 'Respondent id number', 'General happiness'), c('year', 'belief', 'deathPenalty', 'party', 'sex', 'education', 'married', 'workStatus', 'id', 'happiness'))

head(dat)
```

## 1. Find the average years of education for the respondents with each political party affiliation. Arrange your output in a meaningful order and print. Describe any patterns you find.

```{r}
dat$id <- as.numeric(as.character(dat$id))
dat$education <- as.numeric(as.character(dat$education))

indneardem <- dat %>%
  filter(party == "Ind,near dem" & education > 13)
notstrdemocrat <- dat %>%
  filter(party == "Not str democrat" & education > 12)
strongdemocrat <- dat %>%
  filter(party == "Strong democrat" & education > 12)
indnearrep <- dat %>%
  filter(party == "Ind,near rep" & education > 13)
notstrrepublican <- dat %>%
  filter(party == "Not str republican" & education > 13)
strongrepublican <- dat %>%
  filter(party == "Strong republican" & education > 13)
otherparty <- dat %>%
  filter(party == "Other party" & education > 13)
independent <- dat %>%
  filter(party == "Independent" & education > 12)
dontknow <- dat %>%
  filter(party == "Don't know" & education > 11)
noanswer <- dat %>%
  filter(party == "No answer" & education > 12)

mergeCols <- c('year', 'belief', 'deathPenalty', 'party', 'sex', 'education', 'married', 'workStatus', 'id', 'happiness')

MedEduByParty <- data.frame(aggregate(dat$education ~ dat$party, data=dat, median))

MedEduByPartyOrdered <- MedEduByParty[order(MedEduByParty$dat.education), ]

MedEduByPartyOrdered

```

    Those who responded "Don't know" have the lowest median education level, while those who responded "Strong republican" had the highest median education level.
    
## 2. Create a single data frame containing records for the better educated respondents in each political party affiliation. A “better educated” respondent is someone who has strictly more years of education than the median in the same party affiliation. Print the structure of the data frame.

```{r}
dat$education <- as.numeric(as.character(dat$education))

betterEducated <- do.call("rbind", list(indneardem, notstrdemocrat, strongdemocrat, indnearrep, notstrrepublican, strongrepublican, otherparty, independent, dontknow, noanswer))

head(betterEducated)
str(betterEducated)
```

## 3. How is the happiness of a respondent related to his/her marriage status? Define that a person is happy if the response to question is “Very happy” or “Pretty happy”.
```{r}
happy <- dat %>%
  filter(happiness == "Very happy" | happiness == "Pretty happy")

result1 <- data.frame(aggregate(happy$happiness ~ happy$married, data=happy, FUN = function(x){NROW(x)}))
result1

ggplot(result1, aes(x=happy.married, y=happy.happiness)) + xlab('Marital Status') + ylab('Count') + ggtitle('The Effect of Marital Status on Happiness') + geom_bar(stat="identity", fill="orange")

nothappy <- dat %>%
  filter(happiness != "Very happy" & happiness != "Pretty happy")

result2 <- data.frame(aggregate(nothappy$happiness ~ nothappy$married, data=nothappy, FUN = function(x){NROW(x)}))
result2

ggplot(result2, aes(x=nothappy.married, y=nothappy.happiness)) + xlab('Marital Status') + ylab('Count') + ggtitle('The Effect of Marital Status on Unhappiness') + geom_bar(stat="identity", fill='blue')
 
```

    A large proportion of the "happy" respondents are married, indicating marriage increases happiniess.  When we look at the "unhappy" responses, we also notice that the majority of respondents are married. However, the proportions of respondents who are divorced or never married are much higher when comopared to the "happy" plot.
    
## 4. Does party affiliation affect the relationship you found in the last question?
```{r}
partyAffil <- data.frame(aggregate(happy$married == 'Married' ~ happy$party, data=happy, FUN = function(x){NROW(x)}))

partyAffil <- rename(partyAffil,c('happy.party','happy.married.....Married.'), c('party', 'married'))
partyAffil

ggplot(partyAffil, aes(x=party, y=married)) + xlab('Party Affiliation') + ylab('Count') + ggtitle('The Effect of Political Party on Happy Marriage') + geom_bar(stat='identity', fill='brown',) + coord_flip()
```

    When we look closer at this distribution, we can see that democrats have a noticeable edge over republicans in terms of happy marriages.  When you compare each level(strong, not str, indnear) of affiliation by party, democratic affiliations tend to have more happy marriages than the corresponding republican affiliations. This distribution shows there are more happily married democrats than happily married republicans.
    
## 5. How is marrital status related to the education of a respondent? What are possible reasons accounting for the patterns you see?
```{r}
marritalEdu <- data.frame(aggregate(dat$education ~ dat$married, data=dat, median))
marritalEdu

ggplot(dat, aes(x=married, y=education)) + xlab('Marital Status') + ylab('Education Level') + ggtitle('The Effect of Marital Status on Education Level') + geom_boxplot()
```

    Marital status doesn't appear to have a large effect on the education level of a respondent, as the median education levels are fairly close together.  However, we can see that people who did not provide their marital status had the highest median education level of 16.  One possible reason this category has a higher median education level is that there aren't many people in this category.  Since there are no outliers for this category and the range is small, we can assume less people are in this category.  A small group of people has less variation, which ultimately results in a higher median education level.
    
## 6. Explore two more interesting questions, and answer those questions using the GSS data.

# Q1: How is work status related to the happiness of a respondent?
```{r}
happy <- dat %>%
  filter(happiness == "Very happy" | happiness == "Pretty happy")

result3 <- data.frame(aggregate(happy$happiness ~ happy$workStatus, data=happy, FUN = function(x){NROW(x)}))
result3

ggplot(result3, aes(x=happy.workStatus, y=happy.happiness)) + xlab('Work Status') + ylab('Count') + ggtitle('The Effect of Work Status on Happiness') + geom_bar(stat="identity", fill="gold") + coord_flip()

nothappy <- dat %>%
  filter(happiness != "Very happy" & happiness != "Pretty happy")

result4 <- data.frame(aggregate(nothappy$happiness ~ nothappy$workStatus, data=nothappy, FUN = function(x){NROW(x)}))
result4

ggplot(result4, aes(x=nothappy.workStatus, y=nothappy.happiness)) + xlab('Work Status') + ylab('Count') + ggtitle('The Effect of Work Status on Unhappiness') + geom_bar(stat="identity", fill='purple') + coord_flip()
```

    A large proportion of the "happy" respondents are working full time. Most of the "unhappy" respondents are also working full time.  However, the proportions of respondents who are unemployed, laid off or retired are much higher when comopared to the "happy" plot.  Possible reasons why these work statuses are not as happy include econmomic insecurity and lonliness.
    
# Q2: How is education level related to work status?
```{r}
eduWork <- data.frame(aggregate(dat$workStatus ~ dat$education, data=dat, FUN = function(x){NROW(x)}))

ggplot(dat, aes(x=education, y=workStatus)) + xlab('Education Level') + ylab('Work Status') + ggtitle('The Effect of Education on Work Status') + geom_boxplot()
```

    This plot shows the distribution of education level for each work status.  By looking at these boxplots we can see that the "No answer" category again has the highest median education level.  Just like in question 5, there probably aren't many people in this group, which is why the distribution is not normal.  Excluding the "No answer" category, we can see the next highest median education level is 14 for "School."  This makes sense because these people are pursuing a higher education degree.  The lowest IQRs for education level belong to the "Keeping house" and "Other" work statuses.
